Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus

### 2024-10-28 paper/paper.tex
some comments

Without the above structural rules we
arrive at an ordered, linear type system, and to summarize this
particular choice of substructural restrictions: when parsing, characters must
be processed
without modification and in the order in which they were written.

@@

For a grammar $A$, the recursive nature of $A^{*}$ is captured by defining
$A^{*}$ to be an inductive\footnote{As we will soon see, our theory defines
  \emph{indexed} inductive types. We can recover non-indexed inductive types by
  indexing over the unit type.} linear
type, as shown in \cref{fig:kleenestarinductive}. Just as in \cref{fig:fingram}, we can show that particular strings match a Kleene
star. In \cref{fig:kleenestarderivation}, we demonstrate that \stringquote{ab}
matches $a^{*} \otimes b$ through repeated application of the Kleene star constructors.

@@

\steven{TODO : describe elimination principle somewhere. Maybe just in prose?}
\steven{I probably can't get away with the inline definition of $f$ in \cref{fig:kleeneabstractproof}}
\begin{figure}
\begin{mathpar}
  \inferrule
  {
    \inferrule
    {~}
    {x_I : I \vdash \nil : A^*}
    \\
    \inferrule
    {
      \inferrule
      {~}
      {x_1 : A , x_2 : A , y : A* \vdash \cons (x_1 , \cons(x_2 , y)) : A^*}
    }
    {x : A \otimes A , y : A^*  \vdash f := \letin {x_1 \otimes x_2} {x} {\cons(x_1 ,
        \cons(x_2 , y))} : A^*}
  }
  {x : (A \otimes A)^* \vdash \fold(\nil , f) : A^* }
\end{mathpar}
\caption{An example mapping out of abstract grammars}
\label{fig:kleeneabstractproof}
\end{figure}

@@

Our method of representing these automata internally is to provide an \emph{indexed}
inductive type of traces through the machine. An ordered, linear theory is
integral to our design of \theoryabbv ; however, there are some aspects of parsing that
cannot be captured through linear types alone. Often one wants to define a
\emph{family} of linear types varying over a set, which is a primary motivator
for us to make \theoryabbv a linear-non-linear theory. For the NFA $N$, an indexed inductive
type of traces through $N$ is given in \cref{fig:nfainductive}.

The type of traces in $N$ is indexed by two pieces of data: a state in $N$ and a
Boolean. For each state $s : N.\mathsf{state}$ and $acc : \Bool$, we define
$\Trace_{N}~s~acc$ which denotes the
type of traces starting at state $s$ and terminating at an arbitrary state. The Boolean flag $acc$ marks if the trace
ends in either an accepting or a rejecting state. Let us dissect the structure
of this type in more detail for state 1 of $N$.

@@

\begin{figure}
\begin{align*}
\data &~\Trace_{N}~1~(acc : \Bool) : L_{i}~\where\\
      & \mathsf{stop}_{1} : \ltonl {\Trace_{N}~1~\true}  \\
      & \mathsf{1to2} : \forall (acc : \Bool) \to \ltonl {(b \lto \Trace_{N}~2~acc \lto \Trace_{N}~1~acc)} \\
      & \mathsf{1to3} : \forall (acc : \Bool) \to \ltonl {(\Trace_{N}~3~acc \lto \Trace_{N}~1~acc)} \\
\\
\data &~\Trace_{N}~2~(acc : \Bool) : L_{i}~\where\\
      & \mathsf{stop}_{2} : \ltonl {\Trace_{N}~2~\false} \\
      & \mathsf{2to2} : \forall (acc : \Bool) \to \ltonl {(a \lto \Trace_{N}~2~acc \lto \Trace_{N}~2~acc)} \\
      & \mathsf{2to3} : \forall (acc : \Bool) \to \ltonl {(b \lto \Trace_{N}~3~acc \lto \Trace_{N}~2~acc)}  \\
\\
\data &~\Trace_{N}~3~(acc : \Bool) : L_{i}~\where\\
      & \mathsf{stop}_{3} : \ltonl {\Trace_{N}~3~\true} \\
      & \mathsf{3to1} : \forall (acc : \Bool) \to \ltonl {(a \lto \Trace_{N}~1~acc \lto \Trace_{N}~3~acc)} \\
\end{align*}
\caption{$N$ as an indexed inductive type}
\label{fig:nfainductive}
\end{figure}

@@

Next we present the full details of the syntax and typing of \theoryname.
\theoryname has two kinds of types, linear types $A$ which represent formal
grammars, and non-linear types $X$, which act like ordinary types or sets. Both
linear and non-linear types are checked for well-formedness in a context
$\Gamma$ containing \emph{only} non-linear variables $\Gamma = x:X,\ldots$.

@@

In this section we go on a tour of \theoryabbv. We start by presenting its syntax.
In order to avoid a notion of raw-term, we present it in an intrinsically-typed fashion,
where only well-typed terms are meaningful. Next, we show how it admits a model in $\Set^{\String}$
where the linear types have natural interpretation in terms of formal grammars --- for instance,
the tensor corresponds to grammar concatenation. We conclude the section by showing how
classes of grammars correspond to certain sub-type-theories inside \theoryabbv.

@@

The structural judgments are shown in \cref{fig:structjdg}, the typing
well-formedness in \cref{fig:typewf}, and the intuitionistic typing
rules in \cref{fig:inttyping}. These are mostly just as they appear in
\cite{krishnaswami_integrating_2015}. It has two main differences. The
first one it the presence of two-distinct arrow types, one for adding
variables to the beginning of contexts and another one for adding
variables to the end of contexts. This is an adequate change in the
context of grammars, which are inherently non-symmetric. The second
change is the introduction of inductive types, which allows for the definition
of recursive grammars. The treatment of the non-linear
types is standard, so below we focus on the linear syntax.

@@

\subsection{Formal Grammars as Linear Types}
\label{sec:formaltype}

@@

% \pedro{The paragraphs above are a bit circular, what about the following alternative?}
% \steven{I am in favor of cutting what's above, but we haven't introduced the
%   syntax in the main body of the paper yet, even if the figures appeared
%   earlier. So what you write needs to take that into account}
% \pedro{In which case we should introduce it in the previous section. I suggest
%   adding this right after ``and the intuitionistic typing rules in fig 3''}

Now that we have specified the syntactic aspect of \theoryabbv, we must justify
its connections to formal grammars in order to show that it is a sound methodology
for reasoning about parsers. We do this by showing that $\mathbf{Gr}$ is a model
to our type theory.

@@

\paragraph{Universal Type}
\steven{Definable as nullary product}
The universal type $\top$ may be formed in any context.
%
\[ \sem{\top} \gamma w = \{ \ast \}\]
%
Its grammar semantics in set outputs the unit type in $\Set$ for all input strings in all contexts.

>>>>

\max{Is this really true? I don't think it's accurate to say that the ordering restriction on variable usage actually dictates an order in processing variables. And I don't know what you mean by processed without modification.}
Without the above structural rules we
arrive at an ordered, linear type system, and to summarize this
particular choice of substructural restrictions: when parsing, characters must
be processed
without modification and in the order in which they were written.

@@

For a grammar $A$, the recursive nature of $A^{*}$ is captured by defining
$A^{*}$ to be an \max{this footnote seems unnecessary}inductive\footnote{As we will soon see, our theory defines
  \emph{indexed} inductive types. We can recover non-indexed inductive types by
  indexing over the unit type.} linear
type, as shown in \cref{fig:kleenestarinductive}. Just as in \cref{fig:fingram}, we can show that particular strings match a Kleene
star. In \cref{fig:kleenestarderivation}, we demonstrate that \stringquote{ab}
matches $a^{*} \otimes b$ through repeated application of the Kleene star constructors.

@@

\steven{TODO : describe elimination principle somewhere. Maybe just in prose?}
\steven{I probably can't get away with the inline definition of $f$ in \cref{fig:kleeneabstractproof}}
\max{shouldn't the fold be applied to $x$ and shouldn't $f$ be $\lambda x.\lambda y.\cdots$?}
\begin{figure}
\begin{mathpar}
  \inferrule
  {
    \inferrule
    {~}
    {x_I : I \vdash \nil : A^*}
    \\
    \inferrule
    {
      \inferrule
      {~}
      {x_1 : A , x_2 : A , y : A* \vdash \cons (x_1 , \cons(x_2 , y)) : A^*}
    }
    {x : A \otimes A , y : A^*  \vdash f := \letin {x_1 \otimes x_2} {x} {\cons(x_1 ,
        \cons(x_2 , y))} : A^*}
  }
  {x : (A \otimes A)^* \vdash \fold(\nil , f) : A^* }
\end{mathpar}
\caption{An example mapping out of abstract grammars}
\label{fig:kleeneabstractproof}
\end{figure}

@@

Our method of representing these automata internally is to provide an \emph{indexed}
inductive type of traces through the machine. An ordered, linear theory is
integral to our design of \theoryabbv ; however, there are some aspects of parsing that
cannot be captured through linear types\max{not exactly clear what you mean, maybe you mean ``simple'' linear types?} alone. Often one wants to define a
\emph{family} of linear types varying over a set, which is a primary motivator
for us to make \theoryabbv a linear-non-linear theory\max{jargon-y. Be more direct}. For the NFA $N$, an indexed inductive
type of traces through $N$ is given in \cref{fig:nfainductive}.

The type of traces in $N$ is indexed by two pieces of data: a state in $N$ and a
Boolean. For each state $s : N.\mathsf{state}$ and $acc : \Bool$, we define
$\Trace_{N}~s~acc$ which denotes the
type of traces starting at state $s$ and terminating at an arbitrary state. The Boolean flag $acc$ marks if the trace
ends in either an accepting or a rejecting state\max{confusing. One sentence you say arbitrary state and then the next sentence you say it has to match $acc$}. Let us dissect the structure
of this type in more detail for state 1 of $N$.

@@

\max{this notation is inconsistent with Agda syntax. You write $acc : \Bool$ but then in the $\mathsf{stop}$ cases you instantiate $\acc$ with a specific boolean}
\max{is the boolean just overcomplicating stuff here?}
\begin{figure}
\begin{align*}
\data &~\Trace_{N}~1~(acc : \Bool) : L_{i}~\where\\
      & \mathsf{stop}_{1} : \ltonl {\Trace_{N}~1~\true}  \\
      & \mathsf{1to2} : \forall (acc : \Bool) \to \ltonl {(b \lto \Trace_{N}~2~acc \lto \Trace_{N}~1~acc)} \\
      & \mathsf{1to3} : \forall (acc : \Bool) \to \ltonl {(\Trace_{N}~3~acc \lto \Trace_{N}~1~acc)} \\
\\
\data &~\Trace_{N}~2~(acc : \Bool) : L_{i}~\where\\
      & \mathsf{stop}_{2} : \ltonl {\Trace_{N}~2~\false} \\
      & \mathsf{2to2} : \forall (acc : \Bool) \to \ltonl {(a \lto \Trace_{N}~2~acc \lto \Trace_{N}~2~acc)} \\
      & \mathsf{2to3} : \forall (acc : \Bool) \to \ltonl {(b \lto \Trace_{N}~3~acc \lto \Trace_{N}~2~acc)}  \\
\\
\data &~\Trace_{N}~3~(acc : \Bool) : L_{i}~\where\\
      & \mathsf{stop}_{3} : \ltonl {\Trace_{N}~3~\true} \\
      & \mathsf{3to1} : \forall (acc : \Bool) \to \ltonl {(a \lto \Trace_{N}~1~acc \lto \Trace_{N}~3~acc)} \\
\end{align*}
\caption{$N$ as an indexed inductive type}
\label{fig:nfainductive}
\end{figure}

@@

\max{was this paragraph supposed to be deleted?}
Next we present the full details of the syntax and typing of \theoryname.
\theoryname has two kinds of types, linear types $A$ which represent formal
grammars, and non-linear types $X$, which act like ordinary types or sets. Both
linear and non-linear types are checked for well-formedness in a context
$\Gamma$ containing \emph{only} non-linear variables $\Gamma = x:X,\ldots$.

@@

In this section we go on a tour of \theoryabbv. We start by presenting its syntax.
In order to avoid a notion of raw-term, we present it in an intrinsically-typed fashion,
where only well-typed terms are meaningful\max{this sentence says the same thing multiple times}. Next, we show how it admits a model in $\Set^{\String}$
where the linear types have natural interpretation in terms of formal grammars --- for instance,
the tensor corresponds to grammar concatenation. We conclude the section by showing how
classes of grammars correspond to certain sub-type-theories inside \theoryabbv.\max{why are you putting the entire paper into this section?}

@@

The structural judgments are shown in \cref{fig:structjdg}, the typing
well-formedness in \cref{fig:typewf}, and the intuitionistic typing
rules in \cref{fig:inttyping}. These are mostly just as they appear in
\cite{krishnaswami_integrating_2015}. It has two main differences\max{you absolutely cannot just describe the differences between our calculus and one from prior work. This paper needs to be self-contained}. The
first one it the presence of two-distinct arrow types, one for adding
variables to the beginning of contexts and another one for adding
variables to the end of contexts. This is an adequate\max{adequate?} change in the
context of grammars, which are inherently non-symmetric. The second
change is the introduction of inductive types, which allows for the definition
of recursive grammars. The treatment of the non-linear
types is standard, so below we focus on the linear syntax.

@@

\subsection{Formal Grammars as Linear Types}
\label{sec:formaltype}
\max{this should be in its own section on semantics}

@@

% \pedro{The paragraphs above are a bit circular, what about the following alternative?}
% \steven{I am in favor of cutting what's above, but we haven't introduced the
%   syntax in the main body of the paper yet, even if the figures appeared
%   earlier. So what you write needs to take that into account}
% \pedro{In which case we should introduce it in the previous section. I suggest
%   adding this right after ``and the intuitionistic typing rules in fig 3''}
Now that we have specified the syntactic aspect of \theoryabbv, we must justify
its connections to formal grammars in order to show that it is a sound methodology
for reasoning about parsers. We do this by showing that $\mathbf{Gr}$ is a model
to our type theory.

@@

\paragraph{Universal Type}\max{this is absolutely not the right terminology. Universal type means multiple other completely different things}
\steven{Definable as nullary product}
The universal type $\top$ may be formed in any context.
%
\[ \sem{\top} \gamma w = \{ \ast \}\]
%
Its grammar semantics in set outputs the unit type in $\Set$ for all input strings in all contexts.
